---
title: Advent of Code 2021
author: Dusty Turner
date: '2021-12-01'
slug: []
categories:
  - R
  - datascience
  - learning
  - teaching
tags:
  - R
  - Data Science
  - learning
  - teaching
cover: /media/advent2.JPG
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, paged.print=FALSE, eval=F)
```

Every year, the great team at [Advent of Code](adventofcode.com) lead by [ericwastl](https://twitter.com/ericwastl) develops an "Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like."  

Well, I like R, so in this post I will provide my solutions to the puzzles each day.  I plan on using this post to learn and teach and hopefully I am able to stay up-to-date!  

A few notes as we get started:

- Unless otherwise specified, all solutions rely on `library(tidyverse)`.
- All data is saved in the working directory as a `.txt` file.
- I can't promise my solutions will be the most elegant, but I will only post if they are correct!

On to the solutions...

#### Code supporting all solutions

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
```

## [Day 2](https://adventofcode.com/2021/day/2)

#### Part 1

```{r}
tibble(x = read_lines("day_2.txt")) %>% 
  separate(col = x, into = c("direction","distance"), convert = TRUE) %>% 
  group_by(real_direction = direction == "forward") %>% 
  mutate(distance = ifelse(direction == "up", distance * -1, distance)) %>% 
  summarise(movement = sum(distance)) %>% 
  pull(movement) %>% prod()
```

#### Part 2

```{r}
tibble(x = read_lines("day_2.txt")) %>% 
  separate(col = x, into = c("direction","distance"), convert = TRUE) %>%
  mutate(distance = as.double(distance)) %>% 
  mutate(aim_change = case_when(direction == "forward" ~ 0,
                                direction == "down" ~ distance,
                                direction == "up" ~ -1 * distance)) %>% 
  mutate(aim = cumsum(aim_change)) %>% 
  mutate(horizontal_position_increase = ifelse(direction == "forward", distance, 0)) %>% 
  mutate(depth_position_increase = ifelse(direction == "forward", aim * distance,0)) %>% 
  mutate(horizontal_position = cumsum(horizontal_position_increase)) %>% 
  mutate(depth_position = cumsum(depth_position_increase)) %>% 
  slice_tail(n = 1) %>% 
  summarise(answer = horizontal_position * depth_position) 

```


### [Day 1](https://adventofcode.com/2021/day/1)

#### Part 1

```{r}
tibble(measurements = as.numeric(read_lines("day1.txt"))) %>%  
  filter(measurements > lag(measurements,1)) %>% 
  nrow()
```

#### Part 2

```{r}
tibble(measurements = as.numeric(read_lines("day1.txt"))) %>%
  mutate(roll_sum = zoo::rollsum(measurements,3, fill = NA, align = "right")) %>%
  filter(roll_sum > lag(roll_sum,1)) %>%
  nrow()
```

