---
title: Advent of Code 2021
author: Dusty Turner
date: '2021-12-01'
slug: []
categories:
  - R
  - datascience
  - learning
  - teaching
tags:
  - R
  - datascience
  - learning
  - teaching
cover: /images/cover.jpg
---

Every year, the great team at [Advent of Code](adventofcode.com) lead by [ericwastl](https://twitter.com/ericwastl) develops an "Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like."  

Well, I like R, so in this post I will provide my solutions to the puzzles each day.  I plan on using this post to learn and teach and hopefully I am able to stay up-to-date!  

A few notes as we get started:

- Unless otherwise specified, all solutions rely on `library(tidyverse)`.
- All data is saved in the working directory as a `.txt` file.
- I can't promise my solutions will be the most elegant, but I will only post if they are correct!

On to the solutions...

## Day 1

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
```

### [Part 1](https://adventofcode.com/2021/day/1)

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
tibble(measurements = as.numeric(read_lines("day1.txt"))) %>%  
  filter(measurements > lag(measurements,1)) %>% 
  nrow()
```

### [Part 2](https://adventofcode.com/2021/day/2)

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
tibble(measurements = as.numeric(read_lines("day1.txt"))) %>%
  mutate(roll_sum = zoo::rollsum(measurements,3, fill = NA, align = "right")) %>%
  filter(roll_sum > lag(roll_sum,1)) %>%
  nrow()
```

