---
title: 'Lucky $even Strategy on The Price Is Right'
author: Dusty Turner
date: '2026-02-22'
slug: lucky-seven-strategy-on-the-price-is-right
tags:
  - R
  - Price Is Right
  - Strategy
  - Web Scraping
  - Simulation
  - Games
cover: /media/tpir_logo.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

In previous posts, I've looked at [spinning the big wheel](https://www.dustysturner.com/post/wheel-spin-strategy-on-the-price-is-right) and [winning Pay the Rent](https://www.dustysturner.com/post/winning-pay-the-rent-on-the-price-is-right) on The Price Is Right.  This time, let's tackle **Lucky $even** -- a deceptively simple game where a little number sense goes a long way.

# How Lucky $even Works

The contestant plays for a car with a 5-digit price (e.g., \$25,475).  The first digit is revealed for free.  The contestant starts with **\$7** and must guess the remaining four digits one at a time.

After each guess, the contestant loses **\$|guess - actual|** from their money.  If they guess exactly right, they lose nothing.  If they still have at least **\$1** after all four guesses, they use that dollar to buy the car.

The question is: **what should you guess for each digit to maximize your chance of winning?**

```{r, eval=TRUE, echo=FALSE}
# Replace with a working Lucky $even YouTube video ID
blogdown::shortcode("youtube", "1sn_oGbS8TQ")
```

# What We'll Do

- Scrape game-by-game results from the [Price Is Right Wiki](https://priceisright.fandom.com/wiki/Lucky_$even)
- Reconstruct actual car price digits from the money trail
- Examine digit distributions by position (thousands, hundreds, tens, ones)
- Find the optimal guess for each position
- Simulate win rates under different strategies

# Scrape the Data

The Price Is Right Wiki has gallery pages documenting every playing of Lucky $even, with image captions describing each digit guess and the contestant's remaining money.  We'll scrape the 2010s and 2020s pages, covering roughly 168 games.

```{r}
library(tidyverse)
library(lubridate)
library(rvest)
library(scales)
```

```{r scrape, cache=TRUE}
urls <- c(
  "https://priceisright.fandom.com/wiki/Lucky_%24even/Gallery/2010s",
  "https://priceisright.fandom.com/wiki/Lucky_%24even/Gallery/2020s"
)

scrape_page <- function(url) {
  page <- read_html(url)
  content <- page %>% html_element(".mw-parser-output")

  # Get headings and gallery captions in document order
  # Fandom wiki uses .lightbox-caption for gallery image captions
  nodes <- content %>% html_elements(".mw-headline, .lightbox-caption")
  texts <- nodes %>% html_text2()
  classes <- nodes %>% html_attr("class")

  # Classify each node
  node_type <- if_else(str_detect(classes, "mw-headline"), "heading", "caption")

  tibble(text = texts, type = node_type)
}

raw_data <- map_dfr(urls, scrape_page, .id = "page")
raw_data
```

# Parse the Game Data

Each game starts with a heading containing the date and episode number, followed by captions describing each digit guess.  The captions follow predictable patterns:

- *"says 5 and is exactly right"* -- no money lost
- *"says 4 and has \$3 left"* -- money dropped to \$3
- *"says 6 and has lost all of her money"* -- money hit \$0

```{r}
# Assign each caption to its preceding heading
games_raw <- raw_data %>%
  mutate(game_id = cumsum(type == "heading")) %>%
  filter(game_id > 0)

# Extract heading info
headings <- games_raw %>%
  filter(type == "heading") %>%
  distinct(game_id, .keep_all = TRUE) %>%
  mutate(
    date_str = str_extract(text, "\\w+ \\d{1,2}, \\d{4}"),
    date = mdy(date_str),
    episode = str_extract(text, "#(\\w+)\\)?", group = 1)
  ) %>%
  select(game_id, heading = text, date, episode)

# Extract guess captions (must contain "says" + a digit)
captions <- games_raw %>%
  filter(type == "caption", str_detect(text, "says \\d")) %>%
  mutate(
    guess = as.integer(str_extract(text, "says (\\d)", group = 1)),
    exactly_right = str_detect(text, "exactly right"),
    money_left = as.integer(str_extract(text, "\\$(\\d+) left", group = 1)),
    lost_all = str_detect(text, "lost all|lost .* only dollar")
  )

captions %>% select(game_id, text, guess, exactly_right, money_left, lost_all) %>% head(10)
```

Now we track money through each game to compute the cost of each guess.  A key detail: some special episodes used \$8 or \$10 instead of \$7, and one game was for a 4-digit boat price.  We'll focus on standard \$7 games with 4 digit guesses.

```{r}
# Detect starting money from context captions
starting_money_overrides <- games_raw %>%
  filter(type == "caption") %>%
  mutate(
    is_lucky_eight = str_detect(text, regex("Lucky Eight|\\$8 instead", ignore_case = TRUE)),
    is_lucky_ten = str_detect(text, regex("Lucky Ten|\\$10 instead", ignore_case = TRUE))
  ) %>%
  group_by(game_id) %>%
  summarize(
    starting_money = case_when(
      any(is_lucky_ten) ~ 10L,
      any(is_lucky_eight) ~ 8L,
      TRUE ~ 7L
    )
  )

# Compute money trail within each game
compute_money_trail <- function(df, start_money) {
  n <- nrow(df)
  money_before <- integer(n)
  money_after <- integer(n)

  for (i in seq_len(n)) {
    money_before[i] <- if (i == 1) start_money else money_after[i - 1]
    money_after[i] <- case_when(
      df$exactly_right[i] ~ money_before[i],
      df$lost_all[i] ~ 0L,
      !is.na(df$money_left[i]) ~ df$money_left[i],
      TRUE ~ NA_integer_
    )
  }

  df %>% mutate(money_before = money_before, money_after = money_after)
}

game_data <- captions %>%
  group_by(game_id) %>%
  mutate(digit_position = row_number()) %>%
  ungroup() %>%
  left_join(starting_money_overrides, by = "game_id") %>%
  group_by(game_id) %>%
  group_modify(~ compute_money_trail(.x, .x$starting_money[1])) %>%
  ungroup() %>%
  mutate(cost = money_before - money_after)
```

# Reconstruct Actual Digits

Since we know the guess and the cost (\$|guess - actual|), the actual digit is either `guess + cost` or `guess - cost`.  When only one of those falls in 0--9, the answer is unambiguous.  When both are valid, we have a genuine ambiguity.

```{r}
game_data <- game_data %>%
  mutate(
    option_high = guess + cost,
    option_low = guess - cost,
    actual = case_when(
      cost == 0 ~ guess,
      option_high > 9 & option_low >= 0 ~ option_low,
      option_low < 0 & option_high <= 9 ~ option_high,
      option_high <= 9 & option_low >= 0 ~ NA_integer_, # ambiguous
      TRUE ~ NA_integer_
    ),
    ambiguous = cost > 0 & option_high <= 9 & option_low >= 0
  )

# Filter to standard games (starting money = 7, exactly 4 guesses)
standard_games <- game_data %>%
  filter(starting_money == 7) %>%
  group_by(game_id) %>%
  filter(max(digit_position) == 4) %>%
  ungroup()

cat("Total guesses:", nrow(standard_games), "\n")
cat("Unambiguous actual digits:", sum(!is.na(standard_games$actual)), "\n")
cat("Ambiguous cases:", sum(standard_games$ambiguous), "\n")
```

```{r}
# Join heading info for win/loss
game_outcomes <- standard_games %>%
  group_by(game_id) %>%
  summarize(
    final_money = last(money_after),
    won = final_money >= 1,
    .groups = "drop"
  )

cat("Games:", nrow(game_outcomes), "\n")
cat("Wins:", sum(game_outcomes$won),
    "(", percent(mean(game_outcomes$won)), ")\n")
```

# Digit Distributions by Position

The four digits correspond to the **thousands**, **hundreds**, **tens**, and **ones** place of the car price.  Let's look at the distribution of actual digits for each position, using only unambiguous cases.

```{r fig.width=10, fig.height=6}
position_labels <- c("1" = "Thousands", "2" = "Hundreds", "3" = "Tens", "4" = "Ones")

digit_dist <- standard_games %>%
  filter(!is.na(actual)) %>%
  mutate(position_label = factor(position_labels[as.character(digit_position)],
                                  levels = c("Thousands", "Hundreds", "Tens", "Ones")))

ggplot(digit_dist, aes(x = factor(actual))) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  facet_wrap(~ position_label, scales = "free_y") +
  labs(
    title = "Distribution of Actual Digits in Lucky $even",
    subtitle = "Based on unambiguous cases from ~168 games (2010s-2020s)",
    x = "Digit Value",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

For the ambiguous cases, we can include both possibilities weighted at 0.5 each to get a fuller picture:

```{r fig.width=10, fig.height=6}
# Build weighted digit distribution including ambiguous cases
weighted_dist <- bind_rows(
  # Unambiguous cases: weight = 1
  standard_games %>%
    filter(!is.na(actual)) %>%
    transmute(digit_position, actual, weight = 1),
  # Ambiguous cases: both options weighted at 0.5
  standard_games %>%
    filter(ambiguous) %>%
    transmute(digit_position, actual = option_high, weight = 0.5),
  standard_games %>%
    filter(ambiguous) %>%
    transmute(digit_position, actual = option_low, weight = 0.5)
) %>%
  mutate(position_label = factor(position_labels[as.character(digit_position)],
                                  levels = c("Thousands", "Hundreds", "Tens", "Ones")))

ggplot(weighted_dist, aes(x = factor(actual), weight = weight)) +
  geom_bar(fill = "darkgreen", alpha = 0.8) +
  facet_wrap(~ position_label, scales = "free_y") +
  labs(
    title = "Weighted Digit Distribution (Including Ambiguous Cases)",
    subtitle = "Ambiguous cases contribute 0.5 to each possible digit",
    x = "Digit Value",
    y = "Weighted Count"
  ) +
  theme_minimal(base_size = 14)
```

# Optimal Strategy

For each digit position, the optimal guess minimizes the **expected cost** E[|guess - actual|].  This is achieved by guessing the **median** of the actual digit distribution.

Let's compute the expected cost of every possible guess (0-9) at each position:

```{r fig.width=10, fig.height=8}
# Compute expected cost for each guess at each position
expected_costs <- weighted_dist %>%
  group_by(digit_position, actual) %>%
  summarize(prob_weight = sum(weight), .groups = "drop") %>%
  group_by(digit_position) %>%
  mutate(prob = prob_weight / sum(prob_weight)) %>%
  ungroup() %>%
  crossing(guess_candidate = 0:9) %>%
  mutate(cost = abs(guess_candidate - actual) * prob) %>%
  group_by(digit_position, guess_candidate) %>%
  summarize(expected_cost = sum(cost), .groups = "drop") %>%
  mutate(position_label = factor(position_labels[as.character(digit_position)],
                                  levels = c("Thousands", "Hundreds", "Tens", "Ones")))

# Find optimal guess per position
optimal <- expected_costs %>%
  group_by(digit_position, position_label) %>%
  slice_min(expected_cost, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(expected_costs, aes(x = guess_candidate, y = expected_cost)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(data = optimal, aes(x = guess_candidate, y = expected_cost),
             color = "red", size = 4) +
  geom_text(data = optimal, aes(x = guess_candidate, y = expected_cost,
                                 label = paste0("Guess ", guess_candidate)),
            vjust = -1.2, color = "red", fontface = "bold") +
  facet_wrap(~ position_label) +
  scale_x_continuous(breaks = 0:9) +
  labs(
    title = "Expected Cost by Guess Value at Each Position",
    subtitle = "Red dot = optimal guess (minimizes expected dollar loss)",
    x = "Guess",
    y = "Expected Cost ($)"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
# Summary table
optimal %>%
  select(Position = position_label, `Optimal Guess` = guess_candidate,
         `Expected Cost` = expected_cost) %>%
  mutate(`Expected Cost` = round(`Expected Cost`, 2)) %>%
  knitr::kable()
```

The total expected cost across all four guesses using the optimal strategy is:

```{r}
total_optimal_cost <- sum(optimal$expected_cost)
cat("Total expected cost: $", round(total_optimal_cost, 2), "\n")
cat("Expected money remaining: $", round(7 - total_optimal_cost, 2), "\n")
```

# Simulation: Comparing Strategies

Let's put this to the test.  We'll simulate 100,000 games under four strategies and compare win rates:

1. **Optimal** -- guess the cost-minimizing digit at each position
2. **Always 5** -- guess 5 for every digit (the "middle of the road" approach)
3. **Always Mode** -- guess the most common digit at each position
4. **Random** -- guess a random digit each time

```{r}
# Build digit probability distributions per position
digit_probs <- weighted_dist %>%
  group_by(digit_position, actual) %>%
  summarize(w = sum(weight), .groups = "drop") %>%
  group_by(digit_position) %>%
  mutate(prob = w / sum(w)) %>%
  ungroup()

# Ensure all digits 0-9 represented for each position
full_probs <- crossing(digit_position = 1:4, actual = 0:9) %>%
  left_join(digit_probs %>% select(digit_position, actual, prob),
            by = c("digit_position", "actual")) %>%
  replace_na(list(prob = 0)) %>%
  group_by(digit_position) %>%
  mutate(prob = prob / sum(prob)) %>%  # renormalize
  ungroup()

prob_matrix <- full_probs %>%
  pivot_wider(names_from = actual, values_from = prob, names_prefix = "d") %>%
  select(-digit_position) %>%
  as.matrix()

# Mode per position
mode_guesses <- full_probs %>%
  group_by(digit_position) %>%
  slice_max(prob, n = 1, with_ties = FALSE) %>%
  pull(actual)

optimal_guesses <- optimal$guess_candidate

# Simulation function
simulate_strategy <- function(strategy_fn, n_games = 100000) {
  results <- replicate(n_games, {
    money <- 7
    for (pos in 1:4) {
      actual_digit <- sample(0:9, 1, prob = prob_matrix[pos, ])
      guess <- strategy_fn(pos)
      money <- money - abs(guess - actual_digit)
      if (money <= 0) break
    }
    money
  })
  results
}

set.seed(42)
results <- tibble(
  strategy = c("Optimal", "Always 5", "Always Mode", "Random"),
  money = list(
    simulate_strategy(function(pos) optimal_guesses[pos]),
    simulate_strategy(function(pos) 5),
    simulate_strategy(function(pos) mode_guesses[pos]),
    simulate_strategy(function(pos) sample(0:9, 1))
  )
) %>%
  unnest(money) %>%
  mutate(won = money >= 1)
```

```{r fig.width=10, fig.height=6}
# Win rate comparison
win_rates <- results %>%
  group_by(strategy) %>%
  summarize(
    win_rate = mean(won),
    avg_money = mean(money),
    .groups = "drop"
  ) %>%
  mutate(strategy = fct_reorder(strategy, win_rate))

ggplot(win_rates, aes(x = strategy, y = win_rate, fill = strategy)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  geom_text(aes(label = percent(win_rate, accuracy = 0.1)), vjust = -0.5, fontface = "bold") +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Win Rate by Strategy (100,000 Simulated Games)",
    x = "Strategy",
    y = "Win Rate"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
# Distribution of remaining money
ggplot(results %>% filter(strategy %in% c("Optimal", "Always 5")),
       aes(x = money, fill = strategy)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  annotate("text", x = 1.3, y = Inf, label = "Win threshold ($1)",
           vjust = 2, color = "red", fontface = "italic") +
  labs(
    title = "Distribution of Remaining Money",
    subtitle = "Optimal vs. Always-5 Strategy",
    x = "Money Remaining ($)",
    y = "Count",
    fill = "Strategy"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
win_rates %>%
  mutate(
    win_rate = percent(win_rate, accuracy = 0.1),
    avg_money = paste0("$", round(avg_money, 2))
  ) %>%
  arrange(desc(win_rate)) %>%
  rename(Strategy = strategy, `Win Rate` = win_rate, `Avg Money Remaining` = avg_money) %>%
  knitr::kable()
```

# How Does This Compare to Reality?

```{r}
actual_win_rate <- mean(game_outcomes$won)
cat("Actual win rate from wiki data:", percent(actual_win_rate, accuracy = 0.1), "\n")
cat("Optimal strategy win rate:", percent(win_rates$win_rate[win_rates$strategy == "Optimal"]), "\n")
```

Contestants on the show don't play optimally -- they're guessing under pressure without data.  But the actual win rate gives us a nice benchmark.  The gap between reality and optimal represents how much better contestants *could* do with a data-driven approach.

# Conclusion

Lucky \$even rewards contestants who have good intuition about car prices.  But even without knowing the exact price, a data-driven strategy can meaningfully improve your odds:

- **Car price digits are not uniformly distributed.**  Certain digits appear far more often at each position, driven by how manufacturers price their cars.
- **The optimal guess at each position is the median of the historical digit distribution** -- this minimizes the expected dollar loss per guess.
- **Guessing strategically can substantially improve your win rate** over naive approaches like always guessing 5 or picking at random.

So the next time Drew Carey calls you down and you find yourself playing Lucky \$even, remember: the data is on your side.  Just don't forget to bring that last dollar.
